Explore normal cell types in Neuroblastoma samples

### 2025-06-06 analyses/infercnv-consensus-cell-type/exploratory-notebooks/SCPCP000004/01_nb-consensus-cell-types.Rmd

# Path to cell-type-consensus results
consensus_dir <- file.path(result_dir, "cell-type-consensus", "SCPCP000004")
```

@@

# Path to all possible immune consensus cell types
immune_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-immune-cell-types.tsv"
```

We'll also define vectors for the consensus cell types to check for.

```{r}
immune_celltypes <- readr::read_tsv(immune_url) |>
  dplyr::pull(consensus_annotation)

# Vector of all possible endothelial cell types
endo_celltypes <- c(
  "endothelial cell",
  "blood vessel endothelial cell",
  "microvascular endothelial cell",
  "pericyte"
)
```

@@

# Prepare data frame of all cell types
celltype_df <- celltype_files |>
  # consider only non-PDX
  purrr::keep_at(tumor_library_ids) |>
  purrr::map(readr::read_tsv) |>
  purrr::list_rbind() |>
  dplyr::select(library_id, barcodes, consensus_annotation) |>
  dplyr::mutate(
    category = dplyr::case_when(
      consensus_annotation %in% immune_celltypes ~ "immune",
      consensus_annotation %in% endo_celltypes ~ "endo",
      .default = "other"
    )
  )
```

The table below shows counts of endothelial, immune, and all other cells across libraries:

```{r}
category_count_df <- celltype_df |>
  dplyr::count(library_id, category)

# format for viewing
category_count_df |>
  tidyr::pivot_wider(
    names_from = category,
    values_from = n,
    values_fill = 0
  ) |>
  # order columns
  dplyr::select(library_id, endo, immune, other) |>
  # this way we can print the full table at once; RMarkdown will only show 10 rows at once
  knitr::kable()
```

Looks like we've got a lot of normal cells here, and plenty of libraries we can run with an internal reference. 
Since we have a lot of immune and endothelial cells on their own, we don't need to consider a combined reference for this project.

There are also some libraries with a very small number of cells which are presumably lower-quality and should not be included in any reference.

First, we'll find any libraries that have fewer than 1000 cells and exclude them.

```{r}
discard_libraries <- category_count_df |>
  dplyr::group_by(library_id) |>
  dplyr::summarize(total_cells = sum(n)) |>
  dplyr::filter(total_cells < 1000) |>
  dplyr::pull(library_id)

@@

How many cells would be in each reference for `endo` and `immune` groups?

```{r}
category_count_df <- category_count_df |>
  dplyr::filter(
    category != "other", # we dont need these counts anymore
    !(library_id %in% discard_libraries)
  )

total_ref_cells_df <- category_count_df |>
  dplyr::group_by(category) |>
  dplyr::summarize(total_cells = sum(n))
total_ref_cells_df
```
Well, that's quite a lot of cells!
Let's visualize which libraries these cells are spread across.
The barplot below shows the proportion of cells in the given category from each library.

```{r}
category_prop_df <- category_count_df |>
  dplyr::inner_join(total_ref_cells_df) |>
  dplyr::mutate(prop_total_ref_cells = n / total_cells)

ggplot(category_prop_df) +
  aes(x = library_id, y = prop_total_ref_cells, fill = category) +
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))
```
The plurality of endothelial cells are from `SCPCL000125`, and the plurality of immune cells are from both `SCPCL000126` and `SCPCL000127`.
These libraries dominate their respective reference cell categories, but cells are still reasonably spread across the other libraries.

Next, let's look more closely at the counts for cell types within these broad categories of `immune` and `endo`, since we may have enough data here to narrow the reference down further:

```{r}
fine_celltype_df <- celltype_df |>
  dplyr::filter(category != "other") |>
  dplyr::count(consensus_annotation, category) |>
  dplyr::arrange(category)

# kable to print the full table at once
knitr::kable(fine_celltype_df)
```

* Endothelial cells are nearly exclusively plain `endothelial cell`
* The immune cells are pretty spread out, and we could have cell-type specific references created from anything with more than 100 cells:

```{r}
fine_celltype_df |>
  dplyr::filter(category == "immune", n > 100) |>
  dplyr::arrange(desc(n))
```
Dominant cell types are T cells (across several categories of T cell) and macrophages.

Finally, let's see which libraries could have an internal reference.
Generally `inferCNV` recommends at least 100 cells in a reference, but since there are so many normal cells here we can increase this threshold since 100 is really a lower bound.
So, for this project, we'll use 250 as a threshold to find which libraries/internal references are possible:

```{r}
category_count_df |>
  dplyr::filter(n >= 250) |>
  dplyr::group_by(library_id) |>
  dplyr::mutate(reference_options = paste(category, collapse = ", ")) |>
  dplyr::select(library_id, reference_options) |>
  unique() |>
  # kable to print the full table at once
  knitr::kable()
```

Notably, we have previously observed with our analysis of `SCPCP000015` in this module that references with immune cells tend to show more total CNV overlap with tumor cells, since immune cells are funky! 
So, it's nice to see that we enough data to be able to use a robust `endo` reference without needing to rely on `immune`, should similar CNV patterns emerge here.

@@

* We can _definitely_ have `endo` and `immune` pooled references, and the references won't be _strongly_ biased towards coming from a specific library
* Since we have such a high number of cells in each reference cell grouping in the first place, a combined reference (e.g. `endo-immune`) does not seem necessary
* We could also consider a subset of immune cell types for a reference, particularly macrophage and T cell, but a full immune reference seems reasonable to start with
* We have 11 libraries which we could use an internal reference with: 3 of them with `endo`, and _all_ of them with `immune`

>>>>

# Path to cell-type-consensus results
consensus_dir <- file.path(result_dir, "cell-type-consensus", "SCPCP000004")

# Path to reference cell type groupings
reference_dir <- file.path(repository_base, "analyses", "infercnv-consensus-cell-type", "references")
```

@@

# reference cell type groups
reference_file <- file.path(reference_dir, "reference-cell-groups.tsv")
```

@@

# get cell groups
cell_group_df <- readr::read_tsv(reference_file)

# Prepare data frame of all cell types
celltype_df <- celltype_files |>
  # consider only non-PDX
  purrr::keep_at(tumor_library_ids) |>
  purrr::map(readr::read_tsv) |>
  purrr::list_rbind() |>
  dplyr::select(library_id, barcodes, consensus_annotation) |>
  dplyr::left_join(cell_group_df, by = "consensus_annotation")
```

We'll first exclude any libraries that have fewer than 1000 cells from consideration in the reference.

```{r}
discard_libraries <- celltype_df |>
  dplyr::count(library_id) |>
  dplyr::filter(n < 1000) |>
  dplyr::pull(library_id)

@@

```{r}
celltype_df <- celltype_df |>
  dplyr::filter(!(library_id %in% discard_libraries))
```

These cell groupings are present in the neuroblastoma samples we're considering:

```{r}
unique(celltype_df$reference_cell_group)
```

We should remove any cell groups which might correspond to tumor cells, including `stromal`, `neuronal`, and `stem cell`.
The `NA` values here correspond to `Unknown` cells, which we'll also remove.
This means we'll keep groups `endothelial`, `epithelial`, and `immune`.

```{r}
exclude_groups <- c("stromal", "neuronal", "stem cell")
celltype_df <- celltype_df |>
  dplyr::filter(
    !(reference_cell_group %in% exclude_groups),
    consensus_annotation != "Unknown"
  )
```

Let's begin by counting how many cells in each cell group we have per library

```{r}
celltype_count_df <- celltype_df |>
  dplyr::count(library_id, reference_cell_group)

# format for viewing
celltype_count_wide_df <- celltype_count_df |>
  tidyr::pivot_wider(
    names_from = reference_cell_group,
    values_from = n,
    values_fill = 0
  ) |>
  # add column for total normal cells
  dplyr::mutate(
    all_normal = rowSums(dplyr::across(where(is.numeric)))
  ) |>
  # arrange by total normal cells
  dplyr::arrange(desc(all_normal))

# this way we can print the full table at once; RMarkdown will only show 10 rows at once
knitr::kable(celltype_count_wide_df)
```

Virtually all of the normal cell types are `endothelial` and `immune`, with very few `adipocyte` and literally only one `epithelial` cell.

We'll look at this data in barplot form to see, per-library, the relative proportions of each normal cell type group out of all normal cells in the library.

```{r}
celltype_prop_df <- celltype_count_wide_df |>
  # long again, but keep all_normal column for calculating proportion
  tidyr::pivot_longer(
    -c(library_id, all_normal),
    names_to = "reference_cell_group",
    values_to = "n"
  ) |>
  dplyr::mutate(cell_group_prop = n / all_normal)

ggplot(celltype_prop_df) +
  aes(x = library_id, y = cell_group_prop, fill = reference_cell_group) +
  geom_col() +
  labs(y = "Proportion of normal cells in group") +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
  )
```

Based on these numbers, we can consider three normal reference options:

* All normal cells
* ALl immune cells
* All endothelial cells

How many cells would be in each of those pooled references?

```{r}
# update the count data frame without categories we don't need
celltype_count_df <- celltype_count_wide_df |>
  dplyr::select(-adipocyte, -epithelial) |>
  tidyr::pivot_longer(
    -library_id,
    names_to = "reference_cell_group",
    values_to = "n"
  )

celltype_count_df |>
  dplyr::group_by(reference_cell_group) |>
  dplyr::summarize(total_cells = sum(n))
```

Well, that's quite a lot of cells for these pooled references!
A potential concern is that this many cells will incur prohibitive runtimes, so we may want to explore a subsampling strategy.
Any subsampling strategy should aim to retain the relative cell type proportions across libraries, but we can explore that if runtime becomes a problem when we reach that stage.

Finally, let's see which libraries could have an internal reference.
`inferCNV` recommends at least 100 cells in a reference, so which combinations of library/reference are possible here?

```{r}
celltype_count_df |>
  dplyr::filter(n >= 100) |>
  dplyr::group_by(library_id) |>
  dplyr::mutate(reference_options = paste(reference_cell_group, collapse = ", ")) |>
  dplyr::ungroup() |>
  dplyr::select(library_id, reference_options) |>
  unique() |>
  # add row counter so we can easily see how many libraries
  dplyr::mutate(row_num = dplyr::row_number()) |>
  # kable to print the full table at once
  knitr::kable()
```

@@

* To start, we can consider three references: all normal, all immune, and possibly also all endothelial
  * We may need to consider a subsetting strategy for these references if runtimes are too long since there would be >10,000 cells in the immune-containing pooled normal references
* There are 19 libraries we can use internal references with, all with the option for all normal cells and some with the immune-only or endothelial-only options too
